---
title: 'Lab 4: t-test Assumptions and Transformations'
date: 'ST 411/511: Methods of Data Analysis I'
output:
  html_notebook: default
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load the packages we will use in this lab
library(ggplot2)
library(Sleuth3)
```

## Notes

- Remember that you can click "Preview Notebook" to view this document in the Viewer pane. This may be helpful because math notation that I write between "$" will appear formatted correctly.

## Outline

In today's lab, we will 

- Learn a little more R functionality
- See how to add new variables to a data frame
- See how to examine assumptions of the $t$-test
- See how to do a simulation study for confidence intervals

## Some additional R functionality

So far, we've been dealing with existing datasets that come in the form of *data.frames* and that are available in the `Sleuth3` library. As you've seen a *data.frame* is a specific type of R object -- it's a rectangular array of data, where the rows contain observations on different subjects (or sampling units or experimental units), and the columns contain different variables that are measured and/or recorded on those subjects.

Let's look at `ex0222` from the `Sleuth3` package, which contains data on the Armed Forces Qualifying Test (AFQT) score percentiles and component test scores in arithmetic reasoning, word knowledge, paragraph comprehension, and mathematical knowledge for a sample of 1,278 U.S. women and 1,306 U.S. men in 1981.

Remember you can use `?` to pop up the help file for a data set or function in the help pane, which can be useful for seeing column names or function arguments. The `head()` function will print out the first 6 rows of data, which can be useful for getting a sense of the data and viewing the column names. We can look at the number of rows and columns in a data frame by using the `dim()` function, where the output is the number of rows, then the number of columns.

Also note that when you run this code chunk you have two panels that pop up below the chunk. The first is output from the `head()` command and the second is from the `dim()` command. You can click on them to view that output.

```{r}
print(getwd())
```

```{r}
gend <- read.csv("/Users/chikezic/Desktop/OSU/ST411511/Homework/Report/gen_data.csv")
dim(gend)
```

```{r}
ProjectA <- read.csv("/Users/chikezic/Desktop/OSU/ST411511/Homework/Report/Cholesterol_projectA.csv")
ProjectB <- read.csv("/Users/chikezic/Desktop/OSU/ST411511/Homework/Report/Cholesterol_projectB.csv")
View(ProjectA)
View(ProjectB)
```

$H_0: \mu_D = 0$
$H_A: \mu_D > 0$


```{r}
ggplot(ProjectA, aes(x = DiffsA)) +
  geom_histogram() 
```


```{r}
DiffsA <- ProjectA$Before - ProjectA$After8weeks
m <- median(DiffsA)
m
nDiffsA <- length(DiffsA)
nDiffsA

Kstat <- sum(DiffsA > 0)
Kstat
```
```{r}
binom.test(Kstat, n=nDiffsA, p=0.5, alt="greater")
```

