---
title: "ST 411/511 Homework 7"
author: "Chimdi Chikezie"
date: "Winter 2022"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(ggplot2)
```

# Instructions

This assignment is due by 11:59 PM, March 4th on Canvas via Gradescope. **You should submit your assignment as a PDF which you can compile using the provide .Rmd (R Markdown) template.** 

> Note: Create a PDF by either compiling a PDF directly (via LaTeX) or from a Word Doc. Do not submit a PDF of the HTML output as they're cumbersome and difficult to read.

Include your code in your solutions and indicate where the solutions for individual problems are located when uploading into Gradescope (Failure to do so will result in point deductions). You should also write complete, grammatically correct sentences for your solutions.

Once you've completed the assignment, and before you submit it to Gradescope, you should read the document to ensure that (1) The computed values show up in the document, (2) the document "looks nice" (i.e. doesn't have extraneous code/outputs and includes _just_ the essentials), and (3) ensure the document is "easy" to read.

#### Goals:

1. Get hands on experience conducting ANOVA and Kruskal-Wallis tests.
2. Practice using linear combinations of means to answer scientific questions of interest.
3. Learn how to perform different methods of multiple comparison procedures with R.

```{r, include=FALSE}
library(ggplot2)
library(Sleuth3)
```

```{r, include = FALSE}
library(multcomp)
```

\newpage
## Question 1 (9 points) -  from *Sleuth* 5.23

**Was Tyrannosaurus Rex warm-blooded?** Several measurements of the oxygen isotopic composition of bone phosphate in each of 12 bone specimens from a single *Tyrannosaurus rex* skeleton were taken. It is known that the oxygen isotopic composition of vertebrate bone phosphate is related to the body temperature at which the bone forms. Differences in means at different bone sites would indicate non-constant temperatures throughout the body. Minor temperature differences would be expected in warm-blooded animals. Is there evidence that the means are different for the different bones? The data are in `ex0523` in the `Sleuth3` library.

### (a) (2 points) Plot the oxygen isotopic composition for each of the bones using a side-by-side boxplot. Comment on whether or not you think the population means are the same for all 12 bones based on your plot.

```{r}
ggplot(ex0523, aes(x=Bone, y=Oxygen)) + geom_boxplot()
```


I think that the population means are the same for all 12 bones. The reason behind this is that the means are closely related and from the box plot, we cannot see outliers which might cause an obvious variability.


### (b) (2 points) Perform an analysis of variance to test whether or not all the population mean oxygen isotopic compositions are the same in the 12 bone types. State your $p$-value and conclusion of the test. You may use the built-in ANOVA functions in R.

```{r}
anova(lm(Oxygen ~ Bone, data=ex0523))
```

I reject the null hypothesis for the alternative at 5% significance level because the p-value (9.73e-07) is less than alpha.
We have evidence to show that at least two population means of oxygen at different bone sites are not the same.
We cannot conclude if Tyrannosaurus Rex warm-blooded or not because we only know that at least two means are different.


### (c) (2 points) Assess the assumption that the population variances are the same in each group by creating a diagnostic plot using the residuals (See Lecture 18 for help with this). Does this assumption appear to have been met?

```{r}
ggplot(ex0523, aes(x=Bone, y=Oxygen)) + geom_point()
```
The equal population variance assumption is valid.
From the plot, majority of the points are closer to the middle, the points do not form any shape and there are no obvious outliers.


### (d) (3 points) Perform a Kruskal-Wallis test using the `kruskal.test()` function. What do you conclude from this test? Compare your conclusion with your result from the analysis of variance in part (b).

```{r}
kruskal.test(Oxygen ~ Bone, data=ex0523)
```

1. $H_0: \mu_1 = \mu_2 = ... = \mu_n$
   
2. I reject the null hypothesis for the alternative at 5% significance level because the p-value is less than alpha. We do not have enough evidence to conclude that the population distribution center of the bones are the same.

The Anova test will be more preferred because from the diagnostic plot, there was no variability identified.


\newpage
## Question 2 (4 points)

### (a) (2 points) In comparing 10 groups, a researcher notices that the sample mean of group 7 is the largest and the sample mean of group 3 is the smallest. The researcher then decides to test the hypothesis that $\mu_7-\mu_3=0$. Why should a multiple comparison procedure be used even though there is only one comparison being made?

A multiple comparison procedure is used here because multiple comparisons have been made already which helped them determine the groups with the highest and smallest sample means.



### b) (2 points) When choosing coefficients for a contrast, does the choice of $\{C_1, C_2, \ldots, C_I\}$ give a different $t$-statistic than the choice of $\{4C_1, 4C_2, \ldots, 4C_I\}$? Explain why or why not.

It does not give a different different $t$-statistic it does not depend on the specific values for the coefficients we input.



\newpage
## Question 3 (9 points) - Modified from *Sleuth* 6.17

The relative head length (RHL) is measured for adders (a type of snake) on the Swedish mainland and on groups of islands in the Baltic Sea. Relative head length is adjusted for overall body length, determined separately for males and females. The data are below and additionally you know that the pooled estimate of standard deviation of the RHL measurements was 11.72 based on 230 degrees of freedom.

```{r}
adder <- data.frame(Locality = c("Uppsala", "In-Fredeln", "Inre Hammnskar", "Norrpada",
                                 "Karringboskar", "Angskar", "SvenskaHagarna"), 
                    SampleSize = c(21, 34, 20, 25, 7, 82, 48), 
                    meanRHL = c(-6.98, -4.24, -2.79, 2.22, 1.27, 1.88, 4.98))

```

Consider the question: "Is the _average of the mean relative head lengths for snakes on the Swedish mainland_ equal to the _average of the mean relative head lengths for snakes on islands in the Baltic Sea_?" Uppsala is the mainland, and the other six localities refer to islands in the Baltic Sea.


### (a) (3 points) Write out the coefficients for the linear combination you would use to test this question, and state the null hypothesis you would be testing using statistical notation.


Coefficients: $C_1 = 1, C_2 = -1/6, C_3 = -1/6, C_4 = -1/6, C_5 = -1/6, C_6 = -1/6, C_7 = -1/6$


$H_0: \gamma = 0$
$H_0: \gamma \neq 0$

### (b) (4 points) What is the $t$-statistic for testing the hypothesis in part (a)? Please include in your answer your computed values of $g$ and the standard error of $g$.

```{r}
sp2 <-11.72^2
df <- 230
g <- (-6.98) - (1/6)*((-4.24) + (-2.79) + 2.22 + 1.27 + 1.88 + 4.98)
sprintf("g: %f", g)

coe <- ((1)^2)/21 + ((-1/6)^2)/34 + ((-1/6)^2)/20 + ((-1/6)^2)/25 + ((-1/6)^2)/7 + ((-1/6)^2)/82 + ((-1/6)^2)/48
sprintf("coe: %f", coe)

SE <- sqrt(sp2 * coe)
sprintf("SE(g): %f", SE)
t <- (g-0)/SE
sprintf("t-statistic: %f", t)
```

### (c) (2 points) Find the resulting $p$-value and state your conclusion (Be sure to write a _complete_ conclusion for this test including all the relevant details and interpretations).

```{r}
pvalue <- 2 * (1 - pt(abs(t), df))
sprintf("P-value: %f", pvalue)
```

I reject the null hypothesis at 5% significance level because the p-value is less than alpha. We do  not have enough to conclude that the mean relative head lengths for snakes on the Swedish mainland is equal to the mean relative head lengths for snakes on islands in the Baltic Sea.



\newpage
## Question 4 (12 points) - Modified from *Sleuth* 6.21

Reconsider the education and future income data from your last homework (data: `ex0525`). Find $p$-values and 95% confidence intervals for the difference in means for all pairs of education groups in the following ways:

### (a) (2 points) Using the Tukey-Kramer procedure.

```{r}
Educ_Mod <- lm(Income2005 ~ Educ, data = ex0525)
Educ_Tukey <- glht(Educ_Mod, linfct = mcp(Educ = "Tukey"))
summary(Educ_Tukey)
confint(Educ_Tukey)
```


### (b) (2 points) Without adjusting for multiple comparisons.

```{r}
summary(Educ_Tukey, test=adjusted("none"))
confint(Educ_Tukey, calpha = univariate_calpha())
```


### (c) (3 points) What do you notice by comparing these two methods? Discuss the differences in which pairwise comparisons are _significant_, how wide the confidence intervals are, and which confidence intervals contain 0.

If we use only the Fisher's LSD procedure, we would think all of the comparisons were statistically significant at the 0.05 level. But when we used the Tukey-Kramer adjustment, only eight of them were statistically significant.

The confidence intervals of the Tukey-Kramer are wider unlike that of Fisher's LSD which does not make any adjustment for the multiple comparisons. The point estimates are the same (as we expect), but the lower and upper bounds are changing, showing that the half-widths of the confidence intervals are changing.

The confidence intervals that contain 0 are:
1. <12 - 12
2. 12 - 13-15
3. <12 - 16

### (d) (3 points) Use the Dunnett procedure to compare every other group to the group with 12 years of education. Look at both the $p$-values and confidence intervals. Which group means apparently differ from the mean for those with 12 years of education?

```{r}
levels(ex0525$Educ) # print levels before re-leveling
ex0525$Educ <- relevel(ex0525$Educ, ref = "12")
levels(ex0525$Educ) # print levels after re-leveling
Educ_Mod <- lm(Income2005 ~ Educ, data = ex0525)



Educ_Dunnett <- glht(Educ_Mod, linfct = mcp(Educ = "Dunnett"))
summary(Educ_Dunnett)
confint(Educ_Dunnett)
```

The group means that apparently differ from the mean for those with 12 years of education are the means of groups 13-15, 16 and >16 respectively.


### (e) (2 points) Taking all of these tests together, what general statements would you make about the relationship between Education and Income?

The relationship between Education and Income from the analysis which shows some evidence that IQ does not generally determine a person's Future Income.