---
title: "ST 411/511 Homework 5"
author: "Chimdi Chikezie"
date: "Winter 2022"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(ggplot2)
```

# Instructions

This assignment is due by 11:59 PM, February 18th on Canvas via Gradescope. **You should submit your assignment as a PDF which you can compile using the provide .Rmd (R Markdown) template.** 

> Note: Create a PDF by either compiling a PDF directly (via LaTeX) or from a Word Doc. Do not submit a PDF of the HTML output as they're cumbersome and difficult to read.

Include your code in your solutions and indicate where the solutions for individual problems are located when uploading into Gradescope (Failure to do so will result in point deductions). You should also write complete, grammatically correct sentences for your solutions.

Once you've completed the assignment, and before you submit it to Gradescope, you should read the document to ensure that (1) The computed values show up in the document, (2) the document "looks nice" (i.e. doesn't have extraneous code/outputs and includes _just_ the essentials), and (3) ensure the document is "easy" to read.

#### Goals:

1. Practice using, what we often think of as, alternatives to the t-tests.
2. Familiarize ourselves with how the underlying hypotheses are different from the t-test, and amongst the alternatives, to better understand what our conclusions mean.

```{r, include=FALSE}
# Load libraries we need for the assignment
library(ggplot2)
library(Sleuth3)
```


\newpage
## Question 1 (14 points): One Sample Tests
Consider the following data ($n=18$ observations) (Assume $\alpha = 0.05$ unless otherwise specified)

```{r}
data1 <- c(5.4, -13.2, -4.3, 1.3, -14.3, -12.3, -11.4, 9.1, 2.5, 
           -6.8, 2.6, 6.2, 10.4, -17.0, -17.1, -8.0, 16.5, -15.3)
```

### (a) (5 points) Use the sign test to test the null hypothesis that the population median is 0 ($H_0: m=0$) using the following steps:
- State the hypotheses you are testing.
- Compute the sign test statistic $K$ by hand.
- Compute the $z$-statistic corresponding to the value of $K$ that you obtained.
- Find the two-sided $p$-value using the normal approximation to the null distribution of $K$.
- Write a conclusion for your test.

$H_0: m=0$
$H_A: m\neq0$

```{r}
Kstat <- sum(data1 > 0)
Kstat
```

```{r}
n <- 18
z <- (Kstat - (n/2))/sqrt(n/4)
z
```

```{r}
pvalue <- 2 * (1 - pnorm(abs(z)))
pvalue
```

We fail to reject the null hypothesis for the alternative hypothesis at 5% significance level because the p-value is greater than alpha.
We do not have enough evidence to show that the population median cannot be equal to zero.


\newpage
### (b) (6 points) Use the Wilcoxon Signed-Rank test to test that the "center" of this distribution is 0 using the following steps:
- State the hypotheses you are testing.
- Compute the signed-rank test statistic $S$ by hand.
- Compute the $z$-statistic corresponding to the value of $S$ that you obtained.
- Find the two-sided $p$-value using the normal approximation to the null distribution of $S$.
- Verify that you get the same result by using the `wilcox.test()` function in R with `exact=FALSE` and `correct=FALSE`.
- Write a conclusion for your test.


$H_0: m = 0$
$H_A: m \neq 0$

```{r}
n <- length(data1)

mu0 <- 0
distRanks <- rank(abs(data1 - mu0))

Sstat <- sum(distRanks[data1 > mu0])
Sstat
```

```{r}
z <- ((Sstat - (n*(n + 1) / 4)) / (sqrt((n*(n+1) * (2*n + 1))/24)))
z

```

```{r}

pvalue <- 2 * (1 - pnorm(abs(z)))
sprintf("P-value: %f", pvalue)

wilcox.test(data1, mu=0, alternative = "two.sided", exact = FALSE, correct = FALSE)
```

We fail to reject the null hypothesis at 5% significance level because the p-value is greater than alpha.
We do not have enough evidence to show that center of the population distribution cannot be zero.
Therefore, 0 is a plausible value for the population pseudomedian.


\newpage
### (c) (3 points) Discuss the differences between the results in parts (a) and (b). Specifically, address:
- Do both tests come to similar conclusions (i.e. both reject or both fail to reject)?
- What are the differences between the set of hypotheses, (or are they testing the same thing)? 
- In general, describe why you might choose to use one test versus the other (You might consider how you would explain the results of the test to someone who is not well versed in statistics).


- They both are failing to reject the null hypothesis.
- The sign test is testing for the quantiles or difference in quantiles of a distribution while the Wilcoxon signed rank test is testing for the Pseudomedian (median of all the midpoints).
- Sign test is more understandable because we are testing fot the population median while the Wilcoxon test is testing for center of the population distribution.


\newpage
## Question 2 (16 points): Two-Sample Tests
Consider the guinea pig lifetime data we looked at in Homework 3 (`ex0211` in the `Sleuth3` package). Previously, we considered an equal variance two-sample $t$-test. Now we will consider some alternatives.

### (a) (4 points) Perform a Wilcoxon rank-sum test to test whether the population distributions of lifetime in the Control and Bacilli groups are the same using the `wilcox.text()` function. Consider a one-sided lesser (Bacilli minus Control) alternative hypothesis. You do not need to use the continuity correction. What do you conclude at significance level $\alpha=0.01$?

```{r}
ggplot(ex0211, aes(x = Lifetime)) +
  geom_histogram(binwidth = 40, color="white") +
  facet_wrap(~ Group)
```

H_0: There is no shift
$H_0: P(X>Y) = 0.5$
$H_A: P(X>Y) < 0.5$

```{r}
wilcox.test(Lifetime ~ Group, data = ex0211, alternative="less")
```
I fail to reject the null hypothesis at 0.01 significance level because the p-value is less than alpha.
We do not have enough evidence to say that the population distributions of lifetime in the Control and Bacilli groups are the not same.


\newpage
### (b) (4 points) Perform a Welch's two-sample $t$-test to test the null hypothesis that the population means of lifetime in the Control and Bacilli groups are the same using the `t.test()` function. Consider a one-sided lesser (Bacilli minus Control) alternative hypothesis. What do you conclude at significance level $\alpha=0.01$?

$H_0: \mu_B - \mu_C=0$
$H_A: \mu_B - \mu_C\neq0$

```{r}
t.test(Lifetime ~ Group, data = ex0211, alternative = "less", conf.level = 0.99)

t.test(Lifetime ~ Group, data = ex0211, alternative = "two.sided", conf.level = 0.99)$conf.int
```

Reject the null hypothesis for the alternative hypothese at 0.01 significance level because the p-value is less than alpha.
This means that the population means of lifetime in the Control and Bacilli groups are the not the same.

\newpage
### (c) (3 points) Discuss the differences between the results in parts (a) and (b). Specifically, address:
- Do both tests come to similar conclusions (i.e. both reject or both fail to reject)?
- What are the differences between the set of hypotheses, (or are they testing the same thing)? 
- For this particular experiment, describe why you might choose to use one test versus the other.

-They do not come to similar conclusions.
-The Wilcoxon rank-sum test is testing for an additive effect while Welch Two Sample t-test is testing for the difference in mean of the two groups.
-I will go for the Wilcoxon rank-sum test because our distribution has a lot of outliers.




### (d) (5 points) Previously, in Homework 3, we analyzed the `ex0211` data using the equal variance $t$-test. Create a faceted histogram of the `Lifetime` variable where the facets are based on the `Group` variable (Here, you should consider arranging the facets so that one histogram is on top of the other to help you better compare the difference in the amount of spread. To do this in `ggplot2`, include the `ncol = 1` argument within the `facet_wrap()` function). Based on the sample distributions of each group, argue why we should use Welch's $t$-test instead of the equal variance $t$-test. Then, compare the R output for Welch's $t$-test and the equal variance $t$-test and comment about how the two tests differ in terms of their computed test statistics, degrees of freedom, and $p$-values.


```{r}
ggplot(ex0211, aes(x=Lifetime)) + geom_histogram(color="white") + facet_wrap(~Group, ncol = 1)
```

Welch's t-test will be used because the two distributions have different variances which are distinct, and the Welch's t-test is more resistant to outliers compared to the equal variance t-test.
```{r}
t.test(Lifetime~Group, data = ex0211, alternative = "less", var.equal = TRUE, conf.level=0.99)
```

```{r}
t.test(Lifetime~Group, data = ex0211, alternative = "less", conf.level=0.99)
```

- The two sample t-test is used when the distributions have equal variances while the welch's t-test is useful for conducting hypothesis between two    independent populations when the population variances cannot be assumed to be equal.
- The two sample t-test is not resistant to outliers while the welch's t-test is more resistant to outliers.
