---
title: "ST 411/511 Homework 2"
author: "Chimdi Chikezie"
date: "Winter 2022"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(ggplot2)
```

# Instructions

This assignment is due by 11:59 PM, January 21st on Canvas via Gradescope. **You should submit your assignment as a PDF which you can compile using the provide .Rmd (R Markdown) template.** 

> Note: Create a PDF by either compiling a PDF directly (via LaTeX) or from a Word Doc. Do not submit a PDF of the HTML output as they're cumbersome and difficult to read.

Include your code in your solutions and indicate where the solutions for individual problems are located when uploading into Gradescope (Failure to do so will result in point deductions). You should also write complete, grammatically correct sentences for your solutions.

Once you've completed the assignment, and before you submit it to Gradescope, you should read the document to ensure that (1) The computed values show up in the document, (2) the document "looks nice" (i.e. doesn't have extraneous code/outputs and includes _just_ the essentials), and (3) ensure the document is "easy" to read.

#### Goals:

1. Learn how to conduct $Z$ tests for testing claims about unknown population means.
2. Practice writing conclusions for hypothesis tests that are both statistically rigorous but also convey information to non-statistical audiences.
3. Learn how to compute and interpret confidence intervals for unknown population means.
4. Practice determining whether a study is an observational study or an experiment. 
5. Based on the design of a study, determine the largest population for which the outcome of the study applies.


\newpage
## Question 1 (12 points)

A random sample of $n = 500$ books is selected from a library and the number of words in the title of each book is recorded. The sample mean number of words in the title is 6.2 words. The population variance is 40 words-squared.

### (a) (2 points) Compute the $z$-statistic for testing the null hypothesis $H_0: \mu = 7$.

```{r}
n <- 500
mu <- 7
x <- 6.2
s <- 40

z <- (x - mu) / sqrt(s/n)
z

```


### (b) (3 points) Perform a level $\alpha=0.1$ test of $H_0: \mu = 7$ vs. the one-sided lesser alternative $H_A: \mu < 7$ by comparing the *computed test statistic* (from part (a)) to the correct *critical value*. Be sure to include a *complete* conclusion for your test which states (1) whether you reject or fail to reject the null hypothesis, (2) the reasoning behind why you reject/fail to reject, and (3) what the conclusion means in terms of the context of the question.

```{r}
qnorm(0.1)
```

a - I reject the null hypothesis.
b - The value of the z-test falls with the rejection rejection from the alpha test.
c - This means that the mean of the 500 books sampled is way less than the mean suggested by the null hypothesis.


### (c) (2 points) What is the one-sided lesser $p$-value for the statistic you computed in part (a)?

```{r}
pnorm(z)
```


### (d) (2 points) What is the two-sided $p$-value for the statistic you computed in part (a)?

```{r}
2 * (1 - pnorm(abs(z)))
```


### (e) (2 points) Construct a 95% confidence interval for the population mean number of words per title. Hint: recall that a 95% confidence interval is formed by the sample mean $\pm 1.96 \times$ standard deviation of the sampling distribution. Write a sentence which communicates the bounds of the confidence interval.

```{r}
#n <- 500
#mu <- 7
#x <- 6.2
#s <- 40

err <- 1.96

upper_limit <- x + (err * sqrt(s/n))
lower_limit <- x - (err * sqrt(s/n))

upper_limit
lower_limit

```

The lower bound is 5.64
The upper bound is 6.75


### (f) (1 point) Based on your confidence interval from part (e), would a level $\alpha=0.05$ two-sided hypothesis test reject or fail to reject the null hypothesis that the population mean is 6.5 words per title? How do you know? Answer this question without conducting the two-sided test with $alpha = 0.05$.

It will fail to reject the null hypothesis because the mean doesn't fall into the rejection regions of the confidence interval.


\newpage
## Question 2 (10 points)
Consider the `rivers` data set in `R`, which is a vector of the lengths (in miles) of 144 "major" rivers in North America, as compiled by the US Geological Survey.

```{r}
data(rivers)
```

### (a) (1 point) What is the length of the longest "major" river in North America? Hint: you can find the maximum of a vector using the `max` function.

```{r}
max(rivers)
```

The length of the longest "major" river in North America is 3710 miles.

### (b) (2 points) Create a *population* histogram of the lengths of the major rivers. Describe the center, shape, and spread of the distribution based solely on the plotted distribution. Note: to use ggplot, the data have to be formatted as a data frame. I have given you the line of code that does this.

```{r}
riversdf <- data.frame(rivers)

ggplot(data = riversdf, aes(rivers)) + 
  geom_histogram() + 
  xlab("Length in Miles") + 
  ggtitle("Length of Major Rivers in North America")
```

The distribution of the lengths of the 144 "major" rivers in North America is unimodal and skewed to the right.
Because the graph is skewed to the right, the center is located at the mean which is about 600 Miles and the histogram has a wide spread or variability.


### (c) (1 point) Select a random sample of $n = 30$ rivers, using `set.seed(411511)` to make sure you draw the same random sample each time. What is the sample mean? Note: To use `set.seed(411511)`, you need to include this command *before* you draw your random sample.

```{r}
set.seed(411511)
samp1 <- sample(rivers, size=30, replace=FALSE)
mean(samp1)

sampriversdf <- data.frame(samp1)


ggplot(data = sampriversdf, aes(x = samp1)) + 
  geom_histogram() + 
  xlab("Length in Miles") + 
  ggtitle("Length of Major Rivers in North America for Samp1")
```

The mean of samp1 is 661.9 miles.

### (d) (2 points) Compute the test statistic for a $z$-test of $H_0:\mu=600$ versus $H_A:\mu\neq 600$.

```{r}
z <- (661.9 - 600) / sqrt(var(rivers)/30)
z
```

### (e) (2 points) Find the $p$-value corresponding to your test statistic from part (d). Recall that you are using a two-sided alternative hypothesis.

```{r}
2 * (1 - pnorm(abs(z)))
```

### (f) (2 points) What do you conclude from this hypothesis test at the 0.05 significance level? State your conclusion in a few short sentences. Be sure to include a *complete* conclusion for your test which states (1) whether you reject or fail to reject the null hypothesis, (2) the reasoning behind why you reject/fail to reject, and (3) what the conclusion means in terms of the context of the question.

I fail to reject the null hypothesis because the z-value 0.6865 is in between the critical values +1.96 and -1.96.


\newpage

## Question 3 (3 points)

Researchers are curious about how soil type affects plant growth. To study this, they obtain 100 seeds of a particular plant species from a local seed collector. They randomly choose 50 seeds and plant each in a separate pot filled with soil type A. The remaining 50 seeds are each planted in a separate plot filled with soil type B. The plants receive the same care, and at the end of 3 months the height of each plant is measured.

### (a) (1 point) Is this an example of a randomized experiment or an observational study? Justify your answer.

This is a randomized experiment. Reasons being that the researcher randomly choose the seeds and has control over the variables in the research.

### (b) (2 points) What is the largest population for which an inference can be made based on the design of this study? Justify your answer.

100 because the seeds were not gotten through a random method.